From Reversible Programs to \\ Univalent Universes and Back

### 2017-07-05 upi.tex
add some discussion, remove some material that seemed to veer too far off.  Split discussion and conclusion.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion, Related Work, and Conclusion}
\label{sec:conclusion}

@@

By restricting to finite types, two operational semantics emerge:
\begin{enumerate}
\item a (reversible) operational semantics of programs over finite types, and
\item a (reversible) operational semantics of program transformers of such programs.
\end{enumerate}
$\Pi$ programs can be interpreted as generalized permutations, as they work on
structure as well as the underlying skeleton of finite types, aka the natural numbers.
Many of the transformers can be \emph{oriented}, and thus be interpreted as
\emph{optimizations}.  As with traditional languages, there are different aspects
we can optimize for (program size, number of steps to completion, etc); we will
not touch on this further in this paper.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion and Related Work}
\label{sec:discussion}

@@

On can also flip this around, and use the $\lambda$-calculus as the
internal language for Cartesian Closed Categories.  However, as Shulman
explains well in his draft book~\cite{Shullman} on approaching Categorical
Logic via Type Theory, this works for many other kinds of categories.  As
we are interested in \emph{reversibility}, it is most natural to look at
Groupoids.  Thus $\PiTwo$ represents the simplest non-trivial case of
a (reversible) programming language distilled from such ideas.

What is more surprising is how this also turns out to be a sound
and complete language for describing the univalent universe $\bracket{\bt}$.

\section{Conclusion}
\label{sec:conclusion}

What is $\bracket{S^1}$?  Is it useful for programming?  What about $\bracket{\mathbb{N}}$?

What is the ``right'' generalization of $\bracket{-}$ so that we may have all
the usual finite types (such as the ones available in $\Pi$) properly
represented?

\jacques{It is not clear to me that just taking a disjoint union over all the
  types gives the correct generalization.}

Looking at this from the other end: given some ``exotic'' (but finitely
presented) Groupoid $\mathfrak{G}$, is there always a programming language
which is sound and complete for $\mathfrak{G}$ ?
